---
title: "Final Report: HANG TRAN"
output: rmarkdown::github_document
---

# NHANES PROJECT

## Background/Motivation for the Study

Asian population is one of many underrepresented groups in biology and medical research since the majority of studies concentrate on the Caucasian and European groups. Thus, there is a need for a study on the relationship between BMI index and systolic blood pressure and total of cholesterol level in the Asian population.

## References

Varbo, Anette, Marianne Benn, George Davey Smith, Nicholas J. Timpson, Anne Tybjærg-Hansen, and Børge G. Nordestgaard. "Remnant Cholesterol, Low-Density Lipoprotein Cholesterol, and Blood Pressure as Mediators From Obesity to Ischemic Heart Disease." Circulation Research 116, no. 4 (2015): 665--73. <https://doi.org/10.1161/CIRCRESAHA.116.304846>

## Research Question and Hypothesis

The research question is how do the changes in BMI index and average systolic blood pressure affect the mean response in total cholesterol levels in the Asian population from the NHANES dataset?

The hypothesis is the increase in both BMI and systolic blood pressure will cause an increase in the total HDL cholesterol based on historical studies. The positive linear relationship between BMI and average systolic blood pressure and total cholesterol levels indicates that there is likely an association of developing higher HDL cholesterol if people has higher BMI and average systolic blood pressure.

Data Description and Exploratory Data Analysis

The targeted variables are 3 quantitative variables including Body max index in kg/m2, Combined systolic blood pressure reading in mmHg, and Total HDL cholesterol in mmol/L. The sample are the people who identify themselves as Asian in the NHANES dataset. This dataframe is called asian_df and contains 288 observation on 76 variables.

```{r}
library(NHANES) 
library(dplyr) 
asian_df <- filter(NHANES,Race3 == "Asian") #Include only variables self-identify as Asian in a new data frame
```

The explanatory variable (x1 and x2) are the Body max index in kg/m2 and Combined systolic blood pressure reading in mmHg, respectively. The response variable (y) is the Total HDL cholesterol in mmol/L.

```{r}
#Basic summary statistics 
library(mosaic)
mean(asian_df$Age) # Average age of the subject
median(asian_df$Age)

reg1 <- lm(formula = TotChol ~ BMI, data = asian_df) # Linear regression between BMI and total cholesterol
summary(reg1) 

reg2 <- lm(formula = TotChol ~ BPSysAve, data = asian_df) # Linear regression between average systolic BP and total cholesterol
summary(reg2)

reg3 <- lm(asian_df$TotChol~ asian_df$BMI + log(asian_df$BPSysAve)) # Linear regression between average systolic BP and average systolis BP and total cholesterol
summary(reg3) 
```

The mean age is 35.46 years and the median age is 35 years for the subject in the dataframe. We create a regression model (fit a line) between BMI and total cholesterol.

The equation of the fitted line is total cholesterol = 3.844 + 0.037\*BMI. On average, for one kg/m2 increase in BMI, we expect an 0.037mmol/L increase in total cholesterol. We see a positive relation between BMI and the total cholesterol though the strength of the model is very low sine the r2 is 0.03577 so we can only explain 3% of the variability in total cholesterol.

For the second regression model, we create another fitted line between average systolic blood pressure and total cholesterol. The equation of the fitted line is total cholesterol = 4.0624 + 0.006\*Average systolic blood pressure. On average, for one mmHg increase in Average systolic blood pressure, we expect an 0.006mmol/L increase in total cholesterol. We also see a positive relation between average systolic BP and total cholesterol. However, the strength of the model is low since the r2 is less than 1% so we can only explain about 1% of the variability in total cholesterol.

We see that with 2 simple linear regression models above, there is positive relationship in our targeted variables in which the increase in BMI or average systolic BP lead to the increase in total cholesterol. However, the strength of these 2 model is both low. We now have general understanding of the relationship between our targeted variables.

Next, we will create a multiple linear regression using all our targeted variables and do natural log of average systolic BP to fit our model better. The equation is presented below

Total HDL cholesterol = 1.555 + 0.030*BMI + 0.521*average systolic BP. As BMI increase by 1kg/m2, the total HDL cholesterol increase by 0.030mmol/L on average, controlling for average systolic BP. And, as average systolic BP increase by 1%, the total HDL cholesterol increase by 0.521mmol/L, controlling for BMI.

Visualize the relationships in the targeted variables.

```{r}
#Visualization for the targeted variables 

plot(asian_df$BMI, asian_df$TotChol, main = 
       strwrap("Relationship between BMI and Total HDL Cholesterol level", width = 50), 
     xlab = "BMI (kg/m2)", ylab = "Total HDL cholesterol (mmol/L)") +
  abline(reg1, col = "darkred") + 
  theme_minimal()

plot(asian_df$BPSysAve, asian_df$TotChol, main = 
       strwrap("Relationship between Average systolic BP and Total HDL Cholesterol level", width = 50), 
     xlab = "Average systolic blood pressure (mmHg)", 
     ylab = "Total HDL cholesterol (mmol/L)") + 
  abline(reg2, col = "darkred") + 
  theme_minimal()
```

The first plot is the relationship between BMI and total cholesterol. We can see a positive relationship between BMI and the total cholesterol though there is not a perfect fit line in this plot since there are a lot of variables do not follow the regression. The second plot also shows a positive relationship between average systolic BP and the total cholesterol. Similarly, the majority of the point do not follow along the fitted line. The graph do support the hypothesis that the increase in both BMI and average systolic BP will lead to the increase in total cholesterol though the strength of the model is questionable and there is a lot of variability that cannot be explained.

## Analysis

The test used is t-test of regression slope with the null hypothesis Ho and alternative hypothesis Ha. The first predictor variable is BMI with the partial slope, B1 and the second predictor variable is Average systolic blood pressure with the partial slope, B2. We use this test to determine whether there is a linear regression relationship between BMI and Average systolic blood pressure and total HDL cholesterol level

Ho: the change in total HDL cholesterol level cannot be predicted from both BMI and Average systolic blood pressure (B1 = B2 = 0)

Ha: the change in total HDL cholesterol level can be predicted from either BMI or Average systolic blood pressure (either B1 \# 0 or B2 \# 0)

```{r}
reg3 <- lm(asian_df$TotChol~ asian_df$BMI + log(asian_df$BPSysAve)) # Linear regression between average systolic BP and average systolis BP and total cholesterol summary(reg3)

plot(reg3,1) # Residual plot

plot(reg3,2) # Q-Q plot
```

The estimated intercept, Bo is 1.555, estimated partial slope B1, for BMI is 0.030, and estimated partial slope B2, for natural log of average systolic blood pressure is 0.522. The p-value for estimated partial slope of BMI is 0.028 while the p-value for estimated partial slope for natural log of average systolic blood pressure is 0.338. The multiple R-squared is 0.033 and adjusted R-squared is 0.024. The F-statistics is 3.624 on 2 and 210 degree freedom with the p-value of 0.028

Under the standard significance value of 0.05, we can reject the null hypothesis since we have an estimated partial slope for BMI with p-value of 0.028 which is technically smaller than 0.05. We conclude that BMI is a good predictor variable since the its p-value smaller than 0.05 while average systolic blood pressure is not a good predictor variable since its p-value is larger than 0.05. The multiple R-squared states that we can only explain 3.3% of variability in our model thus the strength of our model is very low. The F-statistic is larger than 1 which indicates that either slope in our model is not zero and also the p-value of 0.028 indicating that one of our predictor variable is useful in predicting total HDL cholesterol value.

We can visualize the model with a residual vs. fitted plot and a Q-Q plot to assess our model normality. In the residual vs. fitted plot, the majority of our observation cluster around the horizontal line at zero with the exception of a few indicated extreme values (outliers), little curvature, and approximately equal variance of point above and below the line. For the Q-Q plot, the majority of observation follow the line with some exceptions points indicating our model follow a normal distribution.

This model is statistically significance if we use the significance value of 0.05. However, the model will indeed be rejected if we use the significance value of 0.01. The multiple R-squared is very low and average systolic blood pressure is not a good predictor thus our model is not practical in predicting total HDL cholesterol level. One of the p-value for estimated partial slope is technically smaller than 0.05 but it is not small enough to be consider statistically significance.

## Conclusions

The t-test of regression slope give us useful statistical information about the relationship between response variable and predictors variable. The p-value of each estimated partial slope determine their usefulness as predictor variables. The model support our hypothesis that the increase in both BMI and systolic blood pressure will cause an increase in the total HDL cholesterol based on historical studies. Nevertheless, our model strength is too low with a large amount of variability that cannot be explained. The addition of the average systolic blood pressure with the natural log is not beneficial for predicting the total HDL cholesterol level. The F-statistic only provide that one of our slope is not zero with a p-value not statistically small enough. The model limitation is that our response variable and predictor variables are not independent with the measurement of BMI, average systolic blood pressure, and total HDL cholesterol level conducted on the same individuals, and the existence of extreme values which also decrease our model strength. In conclusion, our model support our research question but the model is not practical or appropriate for predicting our response variable. We should look at all the information we have in the test result and not only rely on the p-value to determine the usefulness of a linear regression model.
